Genetic draft and valley crossing

### 2018-06-03 ms/valley_crossing.tex
edited intro

%\documentclass[rmp,twocolumn]{revtex4}
\documentclass[rmp]{revtex4}
\usepackage[english]{babel}
\usepackage{amssymb,amsfonts,amsmath}
\usepackage{color}
\usepackage{breqn}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\captionsetup{compatibility=false}

@@

@@

@@

@@

The simplest adaptive scenarios in evolution involve the arisal and fixation of successive beneficial mutations.
This appears to be how Darwin thought most complex systems, such as the vertebrate eye, evolved \citep{darwin_1859}, and this assumption underlies much of the field of adaptive dynamics.
However, many adaptations exist that do not appear to have formed by this process.
One class of such adaptations has been termed ``irreducibly complex" because, in their current forms, they cease to function when one or more of their parts is removed.
\citet{muller_1918} described one method by which such adaptations could evolve: components that originally served independent functions might become dependent on each other as parts of a system.
When any such component of a system is removed, the system ceases to function, giving the impression of ``irreducibility".
An alternate method comes from the metaphor of a ``fitness landscape": a wild type individual must traverse a mutational valley--a region of lower fitness--in order to reach a higher fitness peak.
We seek to characterize this ``valley crossing" process, as there is reason to think it may in fact be a common mode of adaptation \citep{trotter_2014}.

Valley crossing in asexual populations is increasingly well understood \citep{weissman_2009}.
If the number of mutants produced every generation is high, the arisal of a complex adaptation that is destined to fix may be guaranteed.
In small populations, wild type individuals may give rise to neutral or deleterious intermediate mutants, which drift to fixation.
A third possibility is for wild type individuals to birth mutant ``bubbles", or transient subpopulations.
These bubbles are ordinarily doomed to extinction due to drift, but a lucky bubble may give rise to a complex adaptation that sweeps to fixation; this process has been referred to as ``tunneling" \citep{iwasa_2004, weissman_2009}.
Different modes of valley crossing were compared to simple sweep models by \citet{ochs_2015}, which found that valley crossing is disfavored at intermediate population sizes (where tunneling is the major mode of crossing) but can be favored in either large or small populations.
Valley crossing in sexually reproducing populations was by \citet{weissman_2010}, which found that the criterion $r < s$, with $r$ the recombination rate and $s$ the selective advantage of the complex adaptation, is the necessary criterion for recombination to be helpful in crossing a valley.
Fairly low recombination rates can bring mutant subpopulations together, increasing the rate at which complex adaptations arise and sweep.
High recombination rates, on the other hand, lower the valley crossing rate by forcing fit individuals to outcross with the less fit wild type.

Previous studies of valley crossing focused primarily on populations where all individuals are equal in fitness except for the focal loci, i.e., the loci at which the individual mutations comprising the complex adaptation segregate.
This is tantamount to assuming that, if there is variation in the background fitness of the population, it is negligible.
In such populations, genetic drift governs the fate of neutral alleles, as well as the behavior of deleterious or beneficial alleles close to frequency zero or one.
Such alleles can be effectively modeled by a diffusion approximation, and the ancestry of a population is described well by the Kingman coalescent.
When population sizes are very large or the fitness variation in the population is substantial, however, the behavior of genetic variation is governed more by genetic \emph{draft} than by drift \citep{gillespie_2000, gillespie_2001, masel_2011, neher_shraiman_2011}.
Whereas drift is the effect of imperfect sampling from generation to generation, draft is the sum total of hitchhiking effects due to the genetic background.
Draft is fundamentally different from drift: for example, drafting populations exhibit genealogies that are better described by an alternative coalescent process such as the Bolthausen-Sznitman coalescent or another Beta coalescent \citep{neher_hallatschek_2013, brunet_2007, schweinsberg_2017}, and allele frequencies experience large jumps that cannot be encapsulated by a diffusion approximation \citep{neher_shraiman_2011}.
In addition, the frequency spectrum of neutral alleles is non-monotonic, with a marked uptick near frequency 1--or, equivalently, a depression in intermediate frequency alleles, which are swept out quickly \citep{kosheleva_2013, neher_shraiman_2011}.
Only a handful of individuals near the ``nose" of the fitness distribution are likely to persist, and they give rise to the bulk of the future population, carrying linked alleles with them.
The parameter that determines whether drift or draft is more important is the product of the population size $N$ and the standard deviation in fitness $\sigma$ \citep{neher_hallatschek_2013}.

In sexual populations, the accuracy of this description depends on the recombination model.
When recombination is ``free", meaning that child genomes are assembled by randomly shuffling those of their parents, there is a chance for new alleles on weak fitness backgrounds to jump onto fitter ones near the nose of the distribution \citep{neher_shraiman_2011}.
Competition between recombination, which breaks up linkage, and adaptation, which creates it, determines whether draft or drift is more important.
When recombination occurs via crossovers, genomes can be partitioned into effectively asexual ``blocks", whose size depends on the per-site crossover rate and the total fitness variation segregating in the population.
If each block harbors some portion $\sigma_b$ of the total fitness standard deviation $\sigma$, then the conditions $N\sigma_b \gg 1$ and $N\sigma_b \ll 1$ correspond to draft and drift dominance, respectively \citep{neher_kessinger_2013}.

It is increasingly realized that draft may play a critical role in shaping genetic diversity, especially in many microbes of interest, where population sizes can be large but ``effective population sizes" are many orders of magnitude smaller \citep{masel_2011}.
The possibility that neutral variants are affected more strongly by selection at linked sites than by genetic drift, even in organisms like \emph{Drosophila}, is one possible explanation for the ``paradox of variation", the fact that genetic diversity and population size often do not scale linearly \citep{gillespie_2000, gillespie_2001, neher_kessinger_2013, corbett-detig_2015}.
Draft likewise has profoundly different effects on the evolution of complex adaptations.
\citet{neher_shraiman_2011} explored the dynamics of stochastic tunneling in rapidly adapting sexual populations.
The primary mode of recombination considered therein is ``free" recombination, in which parents essentially shuffle the entirety of their genomes during reproduction.
We extend this to asexual populations and comment on alternate modes of recombination, such as crossovers.

The complicating factor is that, in an asexual population, the dynamics are governed almost entirely by what happens in the nose of the fitness distribution, and these dynamics can be highly stochastic without the smoothing effect provided by recombination.
We therefore must focus primarily on simulation approaches, as analytical solutions for the behavior of a mutation in the nose of an asexual population are difficult to obtain.
We show that fitness valley crossing occurs at an overall lower rate in rapidly adapting populations.
However, this is consistent with the fact that \emph{all} forms of adaptation are slowed down.
In addition, we confirm that in rapidly adapting population, adaptive events are more likely to be complex adaptations than they are in drifting, neutrally evolving populations: genetic draft maintains the linkage that makes it possible to leap across fitness landscapes rather than adapt primarily by climbing to local peaks.
These observations add to the growing intuition that irreducible complexity may not be a surprising result of the evolutionary process, but rather an expected one.

\section*{Model}

@@

Finally, we examine the ratio $\alpha = \tau_{\mathrm{valley}}/\tau_{\mathrm{sweep}}$, where $\tau_{\mathrm{sweep}}$ is the time required for a particular beneficial mutation to sweep to fixation and $\tau_{\mathrm{valley}}$ is the time required for a valley crossing event to occur.
A higher value of the ratio $\alpha$ means that valley crossing takes much longer than a simple beneficial sweep does.
Though the precise value of this ratio will generally depend on $\delta$, as well as the selection coefficient $s_{\mathrm{valley}}$ of the complex adaptation and $s_{\mathrm{sweep}}$ of the sweeping solo beneficial mutation, a general trend can nonetheless be observed.
In this case, we set $s_{\mathrm{sweep}} = 0.01$, $s_{\mathrm{sweep}} = 0.1$, and $\mu = 10^{-5}$.
For several combinations of parameter values, we find that the ratio $\alpha$ is smaller for high $\sigma$ than for low $\sigma$: see figure \ref{fig:ratios}.
In figure \ref{fig:alpha_ratio}, we consider the behavior of $\alpha(10^{-2})/\alpha(10^{-6})$, a measure of the extent to which rapid adaptation speeds up valley crossing relative to sweeping (lower values indicate that rapid adaptation favors valley crossing).
Overall, we find that increasing $\sigma$ increases the extent to which valley crossing is favored \emph{except} when $N\mu > 1$.
In this case, stochastic tunneling ceases to be the dominant mode of valley crossing.
\citet{weissman_2009} referred to this regime as ``semi-deterministic tunneling": a more or less stable number of intermediate mutant individuals $N\mu/\delta$ persists in the population at low frequencies, which give rise to double mutants at a low but constant rate.
At high $\sigma$, the number of intermediate mutants is likely to fluctuate substantially \citep{cvijovic_2017}, but the average behavior is more or less the same.
This relation more or less obtains in a rapidly adapting population as well, with substantial fluctuations, so $p_{\mathrm{fix}}$ becomes more important in determining the valley crossing rate: since this scales more weakly with $s$ under draft, the overall crossing rate is slowed.

@@

@@

>>>>

\documentclass[rmp]{revtex4}
\usepackage[english]{babel}
\usepackage{amssymb,amsfonts,amsmath}
\usepackage{color}
\usepackage{breqn}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\captionsetup{compatibility=false}

@@

\usepackage{fontspec}

@@

\usepackage{setspace}

@@

\begin{spacing}{1.5}

@@

The simplest adaptive scenarios in evolution involve the arisal and fixation of successive beneficial mutations.
This appears to be how Darwin thought even complex systems, such as the vertebrate eye, evolved \citep{darwin_1859}, and this assumption underlies many models in evolutionary theory including the those from adaptive dynamics \citep{Geritz:Kisdi:1998,Dercole:Rinaldi:2008} and theories of adaptation \citep{Gillespie:1983,Orr:1998,Gillespie:1991}.
Evolution will certainly proceed in this fashion if there always exists a sequence of mutations from an initial genotype to the high fitness genotype where each mutant genotype in the sequence has a higher fitness than the previous one.
If the individual fitness of each genotype is visualized as a surface or landscape where the axes represent alternative alleles at each locus, then smooth landscapes with a single peak ensure that uphill paths exist no matter where a population starts. However, many empirical fitness landscapes are not completely smooth and have multiple peaks \citep[reviewed in][]{Szendro:Schenk:2013,Visser:Krug:2014,Obolski:Ram:2018}.
In such landscapes, wild type individuals may have to traverse a mutational valley--a region of lower fitness--in order to reach a higher fitness peak.
We seek to characterize this ``valley crossing" process in order to better understand the routes adaptation is likely to take on these rugged empirical landscapes \citep[e.g.,][]{Aguilar-Rodriguez:Payne:2017} and to assess the likelihood of valley crossing under more complex scenarios in which it may in fact be common \citep[e.g.][]{trotter_2014}.

Valley crossing in asexual populations is increasingly well understood \citep{weissman_2009}.
If the number of mutants produced every generation is high, which occurs in large populations with high mutation rates, a complex adaptation requiring multiple mutations that is destined to fix will eventually be generated de novo.  
In small populations, neutral or deleterious intermediate mutants generated by wild type individuals can drift to fixation. Then, the further beneficial mutations necessary for the adaptation can fix due to positive selection.
A third possibility occurs for intermediate population sizes where wild type individuals generate transient mutant subpopulations or ``bubbles".
These bubbles are ordinarily doomed to extinction due to drift, but additional mutations in a lucky bubble may generate a complex adaptation that sweeps to fixation; this process has been referred to as ``tunneling" \citep{iwasa_2004, weissman_2009}.
On its own, valley crossing becomes more likely as the population size increases \citep{weissman_2009}. However, recent work suggests that valley crossing relative to the fixation of a simple beneficial mutation is least likely at intermediate population sizes where tunneling occurs \citet{ochs_2015}. This suggests that tunneling may be the most difficult mode of valley crossing. 
% Valley crossing in sexually reproducing populations was by \citet{weissman_2010}, which found that the criterion $r < s$, with $r$ the recombination rate and $s$ the selective advantage of the complex adaptation, is the necessary criterion for recombination to be helpful in crossing a valley.
% Fairly low recombination rates can bring mutant subpopulations together, increasing the rate at which complex adaptations arise and sweep.
% High recombination rates, on the other hand, lower the valley crossing rate by forcing fit individuals to outcross with the less fit wild type.

Previous studies of valley crossing focus primarily on populations where all individuals are equal in fitness except for the focal loci, i.e., the loci at which the individual mutations comprising the complex adaptation segregate.
This is tantamount to assuming that, if there is variation in the background fitness of the population, it is negligible.
In such populations, genetic drift governs the fate of neutral alleles, as well as the behavior of deleterious or beneficial alleles where they are rare, close to frequency zero, or common, close to frequency one.
Alleles whose dynamics are primarily determined by genetic drift can be effectively modeled by a diffusion approximation \citep{Wright:1945,Kimura:1955,Kimura:1957}, and the ancestry of a population is described well by the classic Kingman coalescent \citep{Kingman:1982} where pairs of branches in the genealogy merge back in time until the most recent common ancestor.
When population sizes are very large or the fitness variation in the population is substantial, however, the behavior of genetic variation is governed more by genetic \emph{draft} than by genetic drift \citep{gillespie_2000, gillespie_2001, masel_2011, neher_shraiman_2011}.
Whereas drift is the effect of imperfect sampling from generation to generation, draft is the sum of hitchhiking effects due to selection on sites linked to the focal loci.
Draft is fundamentally different from drift: for example, drafting populations experience large jumps in allele frequencies that cannot be encapsulated by a diffusion approximation \citep{neher_shraiman_2011}
and have genealogies where more than two branches can merge at once and that are better described by alternative coalescent processes such as the Bolthausen-Sznitman coalescent \citep{neher_hallatschek_2013, brunet_2007, schweinsberg_2017}.
Additionally under genetic draft, the frequency spectrum of neutral alleles is non-monotonic, with a marked uptick near frequency one--or, equivalently, a depression in intermediate frequency alleles, which are quickly swept out of the population \citep{neher_shraiman_2011,kosheleva_2013,neher_hallatschek_2013}.
Only a handful of individuals near the ``nose" of the fitness distribution are likely to persist, and they give rise to the bulk of the future population and carry linked alleles with them.
The parameter that determines whether drift or draft is more important is the product of the population size $N$ and the standard deviation in fitness $\sigma$ \citep{neher_hallatschek_2013}.

It is increasingly realized that draft may play a critical role in shaping genetic diversity, especially in many microbial species, where population sizes can be large but ``effective population sizes" are many orders of magnitude smaller \citep{masel_2011}.
The possibility that neutral variants are affected more strongly by selection at linked sites than by genetic drift, even in organisms like \emph{Drosophila}, is one possible explanation for the ``paradox of variation", the fact that genetic diversity and population size often do not scale linearly \citep{gillespie_2000, gillespie_2001, neher_kessinger_2013, corbett-detig_2015}.
Draft likewise may have profound effects on the evolution of complex adaptations. Preliminary evidence for this comes from
\citet{neher_shraiman_2011} who explored how draft affects valley crossing via stochastic tunneling by calculating the mutant bubble size distribution in a rapidly adapting population.
They found that, compared to drift, draft generally makes small and intermediate sized mutant bubbles much rarer but very large bubbles are much more common. Thus, while draft may make valley crossing via tunneling more difficult, the total effect isn't immediately clear.
Here, we extend these previous results by studying in asexual populations how genetic draft affects the rate of crossing fitness valleys across a range of population sizes.

The complicating factor is that, in an asexual population, the dynamics are governed almost entirely by what happens in the nose of the fitness distribution, and these dynamics can be highly stochastic. In previous approaches, these stochastic effects were smoothed due to the presence of recombination \citep{neher_shraiman_2011}.
We therefore must focus primarily on simulation approaches as analytical solutions for the behavior of a mutation in the nose of an asexual population are difficult to obtain.
We show that fitness valley crossing occurs at an overall lower rate in rapidly adapting populations.
However, this is consistent with the fact that \emph{all} forms of adaptation are slowed down.
In addition, we confirm that in rapidly adapting populations, adaptive fixation of alleles are more likely to be complex adaptations involving a fitness valley than they are in populations where genetic drift is the primary force shaping genetic variation at linked sites; essentially, genetic draft maintains the linkage that makes it possible to leap across fitness landscapes rather than adapt primarily by climbing to local peaks.
These observations add to the growing intuition that complex adaptations that involve evolution across fitness valleys may not be a surprising result of the evolutionary process, but rather an expected one.

\section*{Mathematical background}

@@

Finally, we examine the ratio $\alpha = \tau_{\mathrm{valley}}/\tau_{\mathrm{sweep}}$, where $\tau_{\mathrm{sweep}}$ is the time required for a beneficial mutation to sweep to fixation and $\tau_{\mathrm{valley}}$ is the time required for a valley crossing event to occur.
By comparing how $\alpha$ changes as a function of the strength of selection on linked variation, $\sigma$, we can determine how $\sigma$ affects the speed of valley crossing relative to simple beneficial sweeps. Higher (or lower) values of $\alpha$ mean that valley crossing times increase (or decrease) relative to the time required for a simple beneficial sweep.
Though the precise value of this ratio depends on $\delta$ and the selection coefficients $s_{\mathrm{valley}}$ of the complex adaptation and $s_{\mathrm{sweep}}$ of the sweeping solo beneficial mutation, a general trend nonetheless emerges.
In this case, we set $s_{\mathrm{sweep}} = 0.01$, $s_{\mathrm{sweep}} = 0.1$, and $\mu = 10^{-5}$.
For several combinations of parameter values, we find that the ratio $\alpha$ is smaller for high $\sigma$ than for low $\sigma$: see figure \ref{fig:ratios}.
In figure \ref{fig:alpha_ratio}, we consider the behavior of $\alpha(10^{-2})/\alpha(10^{-6})$, a measure of the extent to which rapid adaptation speeds up valley crossing relative to sweeping (lower values indicate that rapid adaptation favors valley crossing).
Overall, we find that increasing $\sigma$ increases the extent to which valley crossing is favored \emph{except} when $N\mu > 1$.
In this case, stochastic tunneling ceases to be the dominant mode of valley crossing.
\citet{weissman_2009} referred to this regime as ``semi-deterministic tunneling": a more or less stable number of intermediate mutant individuals $N\mu/\delta$ persists in the population at low frequencies, which give rise to double mutants at a low but constant rate.
At high $\sigma$, the number of intermediate mutants is likely to fluctuate substantially \citep{cvijovic_2017}, but the average behavior is more or less the same.
This relation more or less obtains in a rapidly adapting population as well, with substantial fluctuations, so $p_{\mathrm{fix}}$ becomes more important in determining the valley crossing rate: since this scales more weakly with $s$ under draft, the overall crossing rate is slowed.

@@

%%% from into but more relevant here
% In sexual populations, the accuracy of this description depends on the recombination model.
% When recombination is ``free", meaning that child genomes are assembled by randomly shuffling those of their parents, there is a chance for new alleles on weak fitness backgrounds to jump onto fitter ones near the nose of the distribution \citep{neher_shraiman_2011}.
% Competition between recombination, which breaks up linkage, and adaptation, which creates it, determines whether draft or drift is more important.
% When recombination occurs via crossovers, genomes can be partitioned into effectively asexual ``blocks", whose size depends on the per-site crossover rate and the total fitness variation segregating in the population.
% If each block harbors some portion $\sigma_b$ of the total fitness standard deviation $\sigma$, then the conditions $N\sigma_b \gg 1$ and $N\sigma_b \ll 1$ correspond to draft and drift dominance, respectively \citep{neher_kessinger_2013}.

@@

\end{spacing}
